---
title: "R Coding Exercise"
---

## Step 1: install packages (output suppressed)
```{r,include = F}
# Setting up all libraries needed for the script
library(tidyverse)

# loading dslabs package to access Gapminder data
library(dslabs)

```

## Step 2: Getting an overview of the Gapminder data
Reviewing the help file, structure, summary, and object type of the Gapminder data
```{r}
#Produce help file for Gapminder data
help(gapminder)

#Look at structure and summary of Gapminder data
str(gapminder)
summary(gapminder)

#Determining the object type of the data
class(gapminder)
```
## Step 3: Creating dataset restricted to African countries
```{r}
#Restricting data to African countries

#Checking what continents are included in the dataset
table(gapminder$continent)
#Restricting to rows where continent is Africa
africadata <- gapminder %>% filter(continent == 'Africa')
#Checking structure and summary of Africa dataset
str(africadata)
summary(africadata)
```

## Step 4: Creating two objects based on African data
obj1 contains 'infant_mortality' and 'life_expectancy'
obj2 contains 'population' and 'life_expectancy'
```{r}
#Creating object 1 with 'infant_mortality' and 'life_expectancy'
obj1 <- africadata %>% select(c(infant_mortality, life_expectancy))
#checking summary and structure of obj1
str(obj1)
summary(obj1)

#Creating object 2 with 'population' and 'life_expectancy'
obj2 <- africadata %>% select(c(population, life_expectancy))
#checking summary and structure of obj2
str(obj2)
summary(obj2)
```
## Step 5: Initial plotting
Two initial relationships with life expectancy were explored, the relationship with infant mortality and population size. These are modeled in the scatterplots below.  
```{r}
#Plotting life expectancy as a function of infant mortality
ggplot(data=obj1, aes(x = infant_mortality, y = life_expectancy)) + geom_point() + 
  labs(title = "Life expectancy as a function of infant mortality")

#Plotting life expectancy as a function of population
ggplot(data=obj2, aes(x = population, y = life_expectancy)) + geom_point() +  scale_x_continuous(trans='log10') + 
  labs(title = "Life expectancy as a function of population", x = "Population (log10 scale)")

```

In the plots above, many points seemed to follow distinct lines. I had a hunch that these were representing country specific trajectories, so I rerun the graphs using color to represent country.

```{r}
#Checking effect of country on plots above by rerunning the graphs using color to represent country.

#Plotting life expectancy as a function of infant mortality
ggplot(data=africadata, aes(x = infant_mortality, y = life_expectancy)) + 
  geom_point(aes(color = country)) + 
  labs(title = "Life expectancy as a function of infant mortality")

#Plotting life expectancy as a function of population
ggplot(data=africadata, aes(x = population, y = life_expectancy)) + 
  geom_point(aes(color = country)) + scale_x_continuous(trans='log10') + 
  labs(title = "Life expectancy as a function of population", x = "Population (log10 scale)")

```

Both graphs, but particularly the graph of life expectancy as a function of population confirmed this. The observed "streaks" are independent countries' life expectancy trajectories.

## Step 6: More data processing - selecting year with most complete data
To determine the level of missingness in infant mortality by year, I looked at both how the missingness was distributed by year and determined the years with the least amount of missingness (i.e., most complete data)
```{r}
#To determine the number of infant mortality data points by year, I filtered the africadata to restrict to only rows with missing infant mortality, grouped by year, and counted the number of rows in each year using n() within the summarize statement

incomp_year <- africadata %>% 
  filter(is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarize(n = n()) 
incomp_year

#Plot of missingness in infant mortality by year
ggplot(incomp_year, aes(x=year, y=n)) + geom_col() + labs(x="Year", y="Number of countries with missing infant mortality data", title="Histogram of missingness on infant mortality data by year")

#I also wanted to print the number of years with the most data. This follows the same process as above, but I restrict to rows with infant mortality data (i.e., where the value is not NA). And, I restrict the final dataset to years with the most rows of infant mortality data
comp_year <- africadata %>% 
  filter(!is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarize(n = n()) %>%
  filter(n == max(n))
comp_year
```

Years 1960-1981 and 2016 all have missing infant mortlity data for at least one country. 
Data is most complete for years 1982-2015. Year 2000 was selected for use in the remaining exercises due to its completeness.


```{r}
#Restricting africadata to only data from the year 2000
yr2000 <- africadata %>%
  filter(year==2000)

#Checking structure and summary of yr2000 dataset
str(yr2000)
summary(yr2000)

```
## Step 7: Plots for year 2000
I reanalyzed the relationships described in Step 5 (relationships between life expectancy and infant mortality and population size) restricteed ot data from the year 2000. 
```{r}
#Plotting life expectancy as a function of infant mortality
ggplot(data=yr2000, aes(x = infant_mortality, y = life_expectancy)) + 
  geom_point() + 
  labs(title = "Life expectancy as a function of infant mortality in 2000")

#Plotting life expectancy as a function of population
ggplot(data=yr2000, aes(x = population, y = life_expectancy)) + 
  geom_point() +  scale_x_continuous(trans='log10') +
  labs(title = "Life expectancy as a function of population in 2000", x = "Population (log10 scale)")
```
The negative association between life expectancy and infant mortality persisted, but there no longer appears to be a relationship between life expectancy and population size. This will be explored statistically in the next Step. 

## Step 8: Simple models to assess relationships between life expectancy and both infant mortality and population size
Simple linear regression was used to fit 2 models, one for each predictor (infant mortality and population size), modeling their effect on life expectancy. 

```{r}
#Fitting first linear model with infant_mortality as a predictor of life expectancy
fit1 <- lm(life_expectancy ~ infant_mortality, data = yr2000)
#Getting model results
summary(fit1)

```
Infant mortality was significantly, negatively associated with life expectancy (P<0.001). 


```{r}
#Fitting second linear model with population as a predictor of life expectancy
fit2 <- lm(life_expectancy ~ population, data = yr2000)
#Getting model results
summary(fit2)

```
Population was not significantly associated with life expectancy (P=0.616).
