---
title: "R Coding Exercise"
---

## Step 1: install packages (output suppressed)

```{r,include = F}
# Setting up all libraries needed for the script
library(tidyverse)

# loading dslabs package to access Gapminder data
library(dslabs)

library(here) # to set directory 
library(dplyr)
install.packages("kableExtra") # this packege is to use to create an aestatic table 
library(knitr)
library(kableExtra)

```

## Step 2: Getting an overview of the Gapminder data

Reviewing the help file, structure, summary, and object type of the Gapminder data

```{r}
#Produce help file for Gapminder data
help(gapminder)

#Look at structure and summary of Gapminder data
str(gapminder)
summary(gapminder)

#Determining the object type of the data
class(gapminder)
```

## Step 3: Creating dataset restricted to African countries

```{r}
#Restricting data to African countries

#Checking what continents are included in the dataset
table(gapminder$continent)
#Restricting to rows where continent is Africa
africadata <- gapminder %>% filter(continent == 'Africa')
#Checking structure and summary of Africa dataset
str(africadata)
summary(africadata)
```

## Step 4: Creating two objects based on African data

obj1 contains 'infant_mortality' and 'life_expectancy' obj2 contains 'population' and 'life_expectancy'

```{r}
#Creating object 1 with 'infant_mortality' and 'life_expectancy'
obj1 <- africadata %>% select(c(infant_mortality, life_expectancy))
#checking summary and structure of obj1
str(obj1)
summary(obj1)

#Creating object 2 with 'population' and 'life_expectancy'
obj2 <- africadata %>% select(c(population, life_expectancy))
#checking summary and structure of obj2
str(obj2)
summary(obj2)
```

## Step 5: Initial plotting

Two initial relationships with life expectancy were explored, the relationship with infant mortality and population size. These are modeled in the scatterplots below.

```{r}
#Plotting life expectancy as a function of infant mortality
ggplot(data=obj1, aes(x = infant_mortality, y = life_expectancy)) + geom_point() + 
  labs(title = "Life expectancy as a function of infant mortality")

#Plotting life expectancy as a function of population
ggplot(data=obj2, aes(x = population, y = life_expectancy)) + geom_point() +  scale_x_continuous(trans='log10') + 
  labs(title = "Life expectancy as a function of population", x = "Population (log10 scale)")

```

In the plots above, many points seemed to follow distinct lines. I had a hunch that these were representing country specific trajectories, so I rerun the graphs using color to represent country.

```{r}
#Checking effect of country on plots above by rerunning the graphs using color to represent country.

#Plotting life expectancy as a function of infant mortality
ggplot(data=africadata, aes(x = infant_mortality, y = life_expectancy)) + 
  geom_point(aes(color = country)) + 
  labs(title = "Life expectancy as a function of infant mortality")

#Plotting life expectancy as a function of population
ggplot(data=africadata, aes(x = population, y = life_expectancy)) + 
  geom_point(aes(color = country)) + scale_x_continuous(trans='log10') + 
  labs(title = "Life expectancy as a function of population", x = "Population (log10 scale)")

```

Both graphs, but particularly the graph of life expectancy as a function of population confirmed this. The observed "streaks" are independent countries' life expectancy trajectories.

## Step 6: More data processing - selecting year with most complete data

To determine the level of missingness in infant mortality by year, I looked at both how the missingness was distributed by year and determined the years with the least amount of missingness (i.e., most complete data)

```{r}
#To determine the number of infant mortality data points by year, I filtered the africadata to restrict to only rows with missing infant mortality, grouped by year, and counted the number of rows in each year using n() within the summarize statement

incomp_year <- africadata %>% 
  filter(is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarize(n = n()) 
incomp_year

#Plot of missingness in infant mortality by year
ggplot(incomp_year, aes(x=year, y=n)) + geom_col() + labs(x="Year", y="Number of countries with missing infant mortality data", title="Histogram of missingness on infant mortality data by year")

#I also wanted to print the number of years with the most data. This follows the same process as above, but I restrict to rows with infant mortality data (i.e., where the value is not NA). And, I restrict the final dataset to years with the most rows of infant mortality data
comp_year <- africadata %>% 
  filter(!is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarize(n = n()) %>%
  filter(n == max(n))
comp_year
```

Years 1960-1981 and 2016 all have missing infant mortlity data for at least one country. Data is most complete for years 1982-2015. Year 2000 was selected for use in the remaining exercises due to its completeness.

```{r}
#Restricting africadata to only data from the year 2000
yr2000 <- africadata %>%
  filter(year==2000)

#Checking structure and summary of yr2000 dataset
str(yr2000)
summary(yr2000)

```

## Step 7: Plots for year 2000

I reanalyzed the relationships described in Step 5 (relationships between life expectancy and infant mortality and population size) restricteed ot data from the year 2000.

```{r}
#Plotting life expectancy as a function of infant mortality
ggplot(data=yr2000, aes(x = infant_mortality, y = life_expectancy)) + 
  geom_point() + 
  labs(title = "Life expectancy as a function of infant mortality in 2000")

#Plotting life expectancy as a function of population
ggplot(data=yr2000, aes(x = population, y = life_expectancy)) + 
  geom_point() +  scale_x_continuous(trans='log10') +
  labs(title = "Life expectancy as a function of population in 2000", x = "Population (log10 scale)")
```

The negative association between life expectancy and infant mortality persisted, but there no longer appears to be a relationship between life expectancy and population size. This will be explored statistically in the next Step.

## Step 8: Simple models to assess relationships between life expectancy and both infant mortality and population size

Simple linear regression was used to fit 2 models, one for each predictor (infant mortality and population size), modeling their effect on life expectancy.

```{r}
#Fitting first linear model with infant_mortality as a predictor of life expectancy
fit1 <- lm(life_expectancy ~ infant_mortality, data = yr2000)
#Getting model results
summary(fit1)

```

Infant mortality was significantly, negatively associated with life expectancy (P\<0.001).

```{r}
#Fitting second linear model with population as a predictor of life expectancy
fit2 <- lm(life_expectancy ~ population, data = yr2000)
#Getting model results
summary(fit2)

```

Population was not significantly associated with life expectancy (P=0.616).

# Part 2

### Contributors

Muhammad Nasir contibutes in this part

# Step 2.1. Overview of the data

```{r}
library(dslabs)
help("mice_weights")


#Look at structure and summary of Mice_weights data 
str(mice_weights)
summary(mice_weights)

#Determining the object type of the data
class(mice_weights)
```

In this project, I use dataset "mice_weights" from dslabs packages. This data contains Body weights, bone density, and percent fat for mice under two diets: chow and high fat. Data provided by Karen Svenson from Jackson Laboratories. Funding to generate these data came from NIH grant P50 GM070683 awarded to Gary Churchill.

There are 7 variables and 780 observations in this data.Variables include body_weight, bone_density, percent_fat, sex, diet, gen, and litter:

Details - body_weight. Body weight in grams at 19 weeks. - bone_density. Body density. - percent_fat. Percent fat. - sex. The sex of the mice. - diet. The diet of the mice: chow or high fat. - gen. These are outbread mice. This variable denotes the generation. - litter. Which of two litters mice belong to.

# Step 2.2. Creating object/ dataset from mice_weights

Now, I am going to create object from the data for further actions (data cleaning and data analysis)

```{r}
mice <- mice_weights 

head(mice) # check the data

str(mice)
```

# Step 2.3. Data cleaning and Processing

In this part, I do data cleaning, elleminating missing values, and creating new variables

```{r}
mice<- na.omit(mice)

summary(mice)
```

```{r}
plot2_1 <- mice %>% 
  ggplot(aes(body_weight, fill = sex)) +
  geom_density(alpha=0.4)+ # to set the tranparant of the color 
  scale_fill_manual( values = c("F"= "green", "M" = "pink"))+ # to set the color of the area
  theme_bw()+ # to create minimalistic style, it changes default plot into white. 
  ggtitle("Figure 2.1. Mice Body Weight based on Sex") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold.italic"))+
  xlab("Body Weight (grams) in 19 weeks")+
  ylab("Density") +
  labs(caption = "Sourse: Jackson Laboratories ") +
  theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1)) # to create frame of the plot. 

plot2_1

figure_file = here("images","Figure 2.1. Mice Body Waight Based on sex.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot2_1) # save the pictures created 
```

Figure 2.1. shows the distribution of body weight of Mice in grams in 19 weeks. It can be seen that Male mice overall had higher body weight compared to female.

```{r}
plot2_2 <-  mice %>% 
  ggplot(aes( x= gen, y=percent_fat, fill = sex)) +
  geom_boxplot(alpha = 0.5)+
  theme_bw()+ # to make the backround minimalistic and convert into white instead of grids 
  scale_fill_manual(values = c("M" = "blue", "F" = "pink")) +  # Set fill color
  scale_color_manual(values = c("M" = "darkblue", "F" = "darkred")) +  # Set border color
  scale_x_discrete(breaks = c(4, 7, 8, 9, 11), # this is to specify the generation
                   labels= c("Generation\n 4", "Generation\n 7 ", "Generation\n 8", "Generation\n 9", "Generation\n 11"))+ # this is to give lable for the generation
  scale_y_continuous()+
  labs(title = "Figure 2.2. Mice Bone Density by Sex and Generation", # create the title
       x = "Generation of The Mice",
       y = "Percent Fat", 
       caption = "Sourse: Jackson Laboratories")+ 
  theme(plot.title = element_text(hjust = 0.5, face = "bold.italic")) + # to adjust the title position, size, and color
   theme(plot.background = element_rect(color = "black", fill = NA, linewidth = 1)) # to create frame of the plot 

 
plot2_2


figure_file = here("images","Figure 2.2. Mice Bone Density by Generation and Sex.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot2_2) # save the pictures created 
```

From figure 2.2. we can see the percentage of fat of mice based on generation and sex. There are different average percent of fat between male and female mice across generation.

Next, I am going to create plot the effect of body fat on bone density.

```{r}
plot2_3 <- ggplot(data = mice, aes(x= body_weight, y= bone_density))+
  geom_point(aes(color = gen)) +
  geom_smooth(method = "loess", se = FALSE)+
  labs(
    subtitle = "(Figure 2.3. Body Weight on Bone Density of the Mice )", 
    y = "Bone Desnity ",
    x = "Body Weight (Grams)",
    caption = "Source: Gapminder Data"
  ) + # to create the name of each axis
  theme(
    plot.background = element_rect(color = "black", size = 1), # Border around the entire plot (including title)
    plot.subtitle = element_text(hjust = 0.5, vjust = -1, size = 12, color = "gray50", face = "bold.italic"), # to adjust the subtitle position, color and size (note, set vjust=185 for printed pictures, for website, set at 100 )
    plot.caption = element_text(hjust = 1, vjust = -13, size = 10, color = "gray40"), # to adjust the subtitle position, color and size, hjust=horizontal and vjust= vertical 
    plot.margin = margin(10, 10, 45, 10)) # (top, right, bottom, left) Increase the bottom margin to create space for the subtitle below the x-axis
print(plot2_3)


figure_file = here("images","Figure 2.3. Body Weight on Bone Density of the Mice.png") # to set up location for the pictures created 
ggsave(filename = figure_file, plot=plot2_3) # save the pictures created 
```

I want to create plot for bone density by diet

```{r}
table2_1  <- mice %>%
  group_by(gen) %>%  # Group data by generation
  summarise(
    avg_body_weight = mean(body_weight, na.rm = TRUE),  # Calculate average body weight
    avg_bone_density = mean(bone_density, na.rm = TRUE),  # Calculate average bone density
    avg_percent_fat = mean(percent_fat, na.rm = TRUE)  # Calculate average percent fat
  ) %>%
  rename(
    Generation= gen, 
    'Average Body Weigth' = avg_body_weight,
    'Average Bone Density' =  avg_bone_density,
    'Average Percent Fat' = avg_percent_fat
  )

table2_1%>%
  kable("html", caption = "Table 2.1. Summary Table of Mice Data by Generation") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE, color = "white", background = "darkblue") %>% # Style gen column
  column_spec(2, background = "lightblue") %>%  # Style avg_body_weight column
  column_spec(3, background = "lightgreen") %>% # Style avg_bone_density column
  column_spec(4, background = "lightyellow") %>% # Style avg_percent_fat column
  row_spec(0, bold = TRUE, color = "white", background = "gray") # Bold header row
table2_1

```

# Step 2.4. Performing Statistical Models

First, I want to perform simple linear model with body weight as a predictor of bone density of mice in the study

```{r}
model_1 <- lm(bone_density ~ body_weight, data = mice)
summary(model_1)
```

From the model, the intercept is 0.398, meaning that the average bone density of mice is 0.398 when the body weight is zero. However, in the real world, it is not possible that mice had 0 gram body weight. The slop tells us the average increase of bone density for every gram in the increase of body weight. There is statistically significant relationship between body weight and bone density with p\<0.0001.

I want to perform Analysis of Variance (ANOVA) to look at the average difference of percent fat for based on diet and generation.

```{r}
model_2 <- lm(percent_fat ~ diet + gen + diet*gen, data = mice)

summary(model_2)
anova(model_2)
```

Conclusion: 1. There is a statistically significant difference in the mean of percent fat between the different diet group with p\<0.0001. 2. There is a statistically significant difference in the mean of percent fat among the different generation with p\<0.0028. 3. There is no significant interaction between diet and generation.
