---
title: "An example exploratory analysis script"
author: "Andreas Handel"
date: "10/28/2021"
output: html_document
---

  
  
This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. 

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```


Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("starter-analysis-exercise","data","processed-data","processeddata2.rds")
#load data
mydata <- readRDS(data_location)
```
Side note: only thing changed here was changing "processeddata.rds" to "processeddata2.rds" to load in Rayleen's data. 




# Data exploration through tables

Showing a bit of code to produce and save a summary table.


```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("starter-analysis-exercise","results", "tables-files", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results/tables` folder. Structure the folders inside `results` such that they make sense for your specific analysis. Provide enough documentation that someone can understand what you are doing and what goes where. `readme.md` files inside each folder are a good idea.


# Data exploration through figures

Histogram plots for the continuous outcomes. 

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("starter-analysis-exercise","results","figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("starter-analysis-exercise","results","figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("starter-analysis-exercise","results","figures","height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("starter-analysis-exercise","results","figures","height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Exercise 2 Part 2 Graphs 

I will create a boxplot in which Emphasis serves as the X variable and Height serves as the Y variable. I will first choose some fun colors for each Emphasis. I will use the ggplot2 package to create this boxplot (specifying geom as geom_boxplot). Using the labs() function, I will rename the axis labels and the title of the boxplot. Then using scale_fill_manual(), I will incorporate the colors I chose earlier. With scale_y_continuous(), I will widen the scope of the y axis. Using theme(), I will put the legend at the bottom and make the axes and title bold and larger in size (as well as center the title). I will save this boxplot as "exercise2-boxplot.png"
```{r}
custom_colors_boxplot <- c("Biostat" = "#9eadee", "DAM" = "#9eebee", "Epi" = "#9eeea4")
  
plot_boxplot <- ggplot(data = mydata, aes(x = Emphasis, y = Height, fill = Emphasis)) + 
  geom_boxplot() +
  labs(x = "Emphasis", y = "Height (in cm) ", title = "Height by Emphasis in PhD Program") +
  scale_fill_manual(values = custom_colors_boxplot) +
  scale_y_continuous(limits = c(120, 200)) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"))

exercise2_boxplot = here("starter-analysis-exercise","results","figures","exercise2-boxplot.png")
ggsave(filename = exercise2_boxplot, plot= plot_boxplot) 
```
Interpretation: From the boxplot above, it appears as though the students in the Biostat Emphasis have the highest average height and have the greatest variation in height. 

I will create a scatterplot with Weight on the X axis and GPA on the Y axis. I will use the ggplot2 package to create this scatterplot (specifying geom as geom_point). Under geom_point(), I will add a color for the points to be and I will increase their size for visibility. Using the labs() function, I will rename the axis labels and the title of the boxplot. Using theme(), I will put the legend at the bottom and make the axes and title bold and larger in size (as well as center the title). I will save this boxplot as "exercise2-scatterplot.png"

```{r}
plot_scatterplot <- ggplot(data = mydata, aes(x = Weight, y = GPA)) +
  geom_point(color = "#9eadee", size = 2.5) +
  labs(title = "Scatterplot of GPA against Weight ", x = "Weight (in kg)", y = "GPA") +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold")) 

exercise2_scatterplot = here("starter-analysis-exercise","results","figures","exercise2-scatterplot.png")
ggsave(filename = exercise2_scatterplot, plot= plot_scatterplot) 
```
Interpretation: It appears as though there is no true correlation between Weight and GPA as the points are all relatively spread out. There is a minimal amount of clustering in the upper left corner, however this likely isn't significant (considering the sample size is n = 9). 

I will not run any statistical tests to assess the correlation coefficient since this was not stated as part of the instructions. 

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.


