---
title: "CDC Data Exercise"
format: html
---

## Data description
I'll be exploring COVID serologic data. This data set includes serologic results from a CDC assay as well as several commercial assays. The data can be found on the [CDC's data website](https://data.cdc.gov/Coronavirus-and-Other-Respiratory-Viruses/Examination-of-SARS-CoV-2-serological-test-results/hhvg-83jq/about_data). In addition to seroreactivity (yes/no) and titers, the dataset also includes information from PCR results, a marker for current COVID infection.

#### Data import

```{r, echo = FALSE, include = FALSE}
library(tidyverse)
library(naniar)
library(procs)
```
First, I import the data and check the level of missingness for each variable.

```{r}
#Reading in the CSV
covid <- read_csv("Examination_of_SARS-CoV-2_serological_test_results_from_multiple_commercial_and_laboratory_platforms_with_an_in-house_serum_panel_20250202.csv")

#Getting an idea of what's in the data set
str(covid)
glimpse(covid)
#Based on this summary, there don't seem to be any weird values
summary(covid)
#There are no missing values
gg_miss_var(covid)
```

I also am choosing to restrict the dataset to variables with information on PCR status, the CDC in-house assay, and the commercial assay from Abbott. There are also variables for both the titer and the log 10 of the titer. Since this information is redundant, I only kept the log transformed variables as these more closely resemble a normal distribution.
```{r}
#Restricting data to variables PCR status, the CDC in-house assay, and the commercial assay from Abbott
covid_mini <- covid %>%
  select(c('PCR confirmed','Abbott Reactivity','In-House CDC ELISA Reactivity','Abbott S/C Values (Log10)', 'In-House CDC ELISA S/T Values (Log10)'))

#Confirming reduction
str(covid_mini)
```



## Exploratory tables - categorical variables

Next, I chose to look at the categorical variables in the dataset: PCR status, Abbott Reactivity, In-House CDC ELISA Reactivity. These tables show the total sample size (N), count (CNT), and percent (PCT) of each category. I also did a crosstab of each of the categorical variables with each other.

```{r}
# Frequency of PCR status
proc_freq(data = covid_mini, tables = `PCR confirmed`)

# Frequency of Abbott Reactivity
proc_freq(data = covid_mini, tables = `Abbott Reactivity`)

# Frequency of In-House CDC ELISA Reactivity
proc_freq(data = covid_mini, tables = `In-House CDC ELISA Reactivity`)

#The proc means function was struggling with the names of the variables, so I created new variables for each of the categorical variables with shorter names
renames <- covid_mini %>%
  mutate(pcr = `PCR confirmed`) %>%
  mutate(cdc = `In-House CDC ELISA Reactivity`) %>%
  mutate(abbott = `Abbott Reactivity`)

#Cross tab of each of the categorical variables
proc_freq(data = renames, tables = v(pcr * cdc, pcr * abbott, cdc * abbott))

```

## Exploratory anaylses - continuous variables
Next, I looked at the distribution of the continous variables: Abbott S/C Values (Log10), In-House CDC ELISA S/T Values (Log10).

For each variable, I created a table that includes the sample size, mean, standard deviation, min, and max and a histogram of the values. Because there appeared to be a bimodal distribution, I performed these steps by seroreactivity status for the relevant assay. I also assessed the distribution of each log titer (from Abbott and CDC) by PCR status.

```{r}
#Producing a table showing the sample size, mean, standard deviation, min, and max for each continuous variable
#Also creating a figure of each variable

#Overall stats - Abbott
proc_means(data = covid_mini, `Abbott S/C Values (Log10)`)

#Stats by seroreactivity status - Abbott
proc_means(data = covid_mini, `Abbott S/C Values (Log10)`, by = `Abbott Reactivity`)

ggplot(covid_mini, aes(x = `Abbott S/C Values (Log10)`, fill = `Abbott Reactivity`, color = `Abbott Reactivity`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Non-reactive" = "maroon", "Reactive" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")


#Overall stats - CDC
proc_means(data = covid_mini, `In-House CDC ELISA S/T Values (Log10)`)

#Stats by seroreactivity status - CDC
proc_means(data = covid_mini, `In-House CDC ELISA S/T Values (Log10)`, by = `In-House CDC ELISA Reactivity`)

ggplot(covid_mini, aes(x = `In-House CDC ELISA S/T Values (Log10)`, fill = `In-House CDC ELISA Reactivity`, color = `In-House CDC ELISA Reactivity`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Non-reactive" = "maroon", "Reactive" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")

#Distribution of log titers by PCR status
#Abbott
#Stats by seroreactivity status - Abbott
proc_means(data = covid_mini, `Abbott S/C Values (Log10)`, by = `PCR confirmed`)

ggplot(covid_mini, aes(x = `Abbott S/C Values (Log10)`, fill = `PCR confirmed`, color = `PCR confirmed`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Neg" = "maroon", "Pos" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")
#CDC
proc_means(data = covid_mini, `In-House CDC ELISA S/T Values (Log10)`, by = `PCR confirmed`)

ggplot(covid_mini, aes(x = `In-House CDC ELISA S/T Values (Log10)`, fill = `PCR confirmed`, color = `PCR confirmed`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Neg" = "maroon", "Pos" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")

          
```

# Part Two: Generating Synthetic Data

This section was contributed by Natalie Cann. 

## Loading Packages 

I will first load packages that will be of use in this exercise. 
```{r}
# Load packages
library(dplyr)
library(purrr)
library(ggplot2)
library(here)
```

## Assessing Correlations from the Original Dataset

I will create a correlation matrix to visualize the relationships between the variables in the original dataset. 
```{r}
# Using numerical values for the categorical variables
covid_numeric <- covid_mini %>%
  mutate(
    pcr = ifelse(`PCR confirmed` == "Pos", 1, 0),
    abbott = ifelse(`Abbott Reactivity` == "Reactive", 1, 0),
    cdc = ifelse(`In-House CDC ELISA Reactivity` == "Reactive", 1, 0)
  ) %>%
  select(pcr, abbott, cdc)

# Compute correlation matrix
cor_matrix <- cor(covid_numeric, method = "pearson")  
print(cor_matrix)

# Visualize correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "color", col = colorRampPalette(c("red", "white", "blue"))(200),
         addCoef.col = "black", tl.col = "black", tl.cex = 1, number.cex = 1)
``` 
The correlation matrix above displays very strong associations between the variables. The correlations between pcr and pcr, abbott and abbott, as well as cdc and cdc are all 1. This is expected as there should be a 100% match when a variable is compared to itself. The correlation between pcr and abbott is 0.95; the correlation between pcr and cdc is 0.95; the correlation between abbot and cdc is 0.96. These high correlation coefficients suggest that the variables are highly associated with each other. I aim to ensure that the correlations between these variables are similar within the synthetic dataset. 

## Summary Statistics

I will run a summary line and count lines to obtain information on means and standards that will help me to create the synthetic dataset. 
```{r}
summary(covid_mini)

mean(covid_mini$`Abbott S/C Values (Log10)`)
sd(covid_mini$`Abbott S/C Values (Log10)`)

mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)`)
sd(covid_mini$`In-House CDC ELISA S/T Values (Log10)`)  # Fixed closing parenthesis

covid_mini %>% filter(`PCR confirmed` == "Pos") %>% count()
covid_mini %>% filter(`Abbott Reactivity` == "Reactive") %>% count()
covid_mini %>% filter(`In-House CDC ELISA Reactivity` == "Reactive") %>% count()
```

I will now calculate the mean and standard deviation of the log-transformed variables. I will calculate the mean and standard deviation for both the reactive and non-reactive groups. 
```{r}
mean(covid_mini$`Abbott S/C Values (Log10)`)
mean(covid_mini$`Abbott S/C Values (Log10)` == "Reactive")
mean(covid_mini$`Abbott S/C Values (Log10)` == "Non-reactive")
sd(covid_mini$`Abbott S/C Values (Log10)`)
mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)`)
mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)` == "Reactive")
mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)` == "Non-reactive")
sd(covid_mini$`In-House CDC ELISA S/T Values (Log10)`)
```

I will now calculate the mean and standard deviation of Abott S/C Values (Log10) for when abbott reactivity is equal to reactive and then non-reactive.
```{r}
# Mean of Abbott S/C Values (Log10) for "Reactive" Abbott Reactivity
mean_reactive_abbott <- mean(covid_mini$`Abbott S/C Values (Log10)`[covid_mini$`Abbott Reactivity` == "Reactive"])

# Mean of Abbott S/C Values (Log10) for "Non-reactive" Abbott Reactivity
mean_non_reactive_abbott <- mean(covid_mini$`Abbott S/C Values (Log10)`[covid_mini$`Abbott Reactivity` == "Non-reactive"])

# Output the results
mean_reactive_abbott
mean_non_reactive_abbott

# sd of Abbott S/C Values (Log10) for "Reactive" Abbott Reactivity
sd_reactive_abbott <- sd(covid_mini$`Abbott S/C Values (Log10)`[covid_mini$`Abbott Reactivity` == "Reactive"])

# sd of Abbott S/C Values (Log10) for "Non-reactive" Abbott Reactivity
sd_non_reactive_abbott <- sd(covid_mini$`Abbott S/C Values (Log10)`[covid_mini$`Abbott Reactivity` == "Non-reactive"])

# Output the results
sd_reactive_abbott
sd_non_reactive_abbott
```
I will now calculate the mean and standard deviation of In-House CDC ELISA S/T Values (Log10) for when In-House CDC ELISA Reactivity is equal to reactive and then non-reactive.
```{r}
# Mean of In-House CDC ELISA S/T Values (Log10) for "Reactive" In-House CDC ELISA Reactivity
mean_reactive_cdc <- mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)`[covid_mini$`In-House CDC ELISA Reactivity` == "Reactive"])

# Mean of In-House CDC ELISA S/T Values (Log10) for "Non-reactive" In-House CDC ELISA Reactivity
mean_non_reactive_cdc <- mean(covid_mini$`In-House CDC ELISA S/T Values (Log10)`[covid_mini$`In-House CDC ELISA Reactivity` == "Non-reactive"])

# Output the results
mean_reactive_cdc
mean_non_reactive_cdc

# sd of In-House CDC ELISA S/T Values (Log10) for "Reactive" In-House CDC ELISA Reactivity
sd_reactive_cdc <- sd(covid_mini$`In-House CDC ELISA S/T Values (Log10)`[covid_mini$`In-House CDC ELISA Reactivity` == "Reactive"])

# sd of In-House CDC ELISA S/T Values (Log10) for "Non-reactive" In-House CDC ELISA Reactivity
sd_non_reactive_cdc <- sd(covid_mini$`In-House CDC ELISA S/T Values (Log10)`[covid_mini$`In-House CDC ELISA Reactivity` == "Non-reactive"])

# Output the results
sd_reactive_cdc
sd_non_reactive_cdc
``` 
## Generating Synthetic Dataset

Next, I will set a seed for reproducibility and specify 204 as the number of observations. 
```{r}
# Setting seed for reproducibility
set.seed(123) 
# Define number of observations 
n_obs <- 204 # same as original dataset
```

I will now create the synthetic dataset. I will use the notes I took from Module 4B: Synthetic Data to help with this. Throughout the process of creating this dataset, I ran into a few errors which I sent through ChatGPT to obtain help. 
```{r}
# Creation of empty data frame with placeholders for variables
covid_synthetic <- data.frame(
  pcr = character(n_obs), 
  abbott = character(n_obs),
  cdc = character(n_obs),
  Abbott_S_C_Values_Log10 = numeric(n_obs),
  In_House_CDC_ELISA_S_T_Values_Log10 = numeric(n_obs)
)

# Variable 1: pcr
covid_synthetic$pcr <- sample(c("Pos", "Neg"), n_obs, replace = TRUE, prob = c(0.43, 0.57)) # probabilities created from counts line above (87/204 pos; 117/204 neg)

# Variable 2: abbott
covid_synthetic$abbott <- sample(c("Reactive", "Non-reactive"), n_obs, replace = TRUE, prob = c(0.41, 0.59)) # probabilities created from counts line above (84/204 reactive; 120/204 non-reactive)

# Variable 3: cdc
covid_synthetic$cdc <- sample(c("Reactive", "Non-reactive"), n_obs, replace = TRUE, prob = c(0.42, 0.58)) # probabilities created from counts line above (86/204 reactive; 118/204 non-reactive)

# Variable 4: Abbott_S_C_Values_Log10
covid_synthetic$Abbott_S_C_Values_Log10 <- unlist(sapply(covid_synthetic$abbott, function(reactivity) {
  if (reactivity == "Reactive") {
    rnorm(1, mean = 0.6649786, sd = 0.1746322) # Mean and sd for Reactive (calculated above)
  } else {
    rnorm(1, mean = -0.7746825, sd = 0.3248742) # Mean and sd for Non-reactive (calculated above)
  }
}))

## ChatGPT Prompt: # Variable 4: Abbott_S_C_Values_Log10   covid_synthetic$Abbott_S_C_Values_Log10 <- rnorm(n_obs, mean = -0.1818809, sd = 0.76081) # mean (-0.1818809) and sd (0.76081) based on original data --- fix this so it shows the mean for reactive (0.6649786) and non-reactive (-0.7746825)
## Error in $<-.data.frame(*tmp*, Abbott_S_C_Values_Log10, value = list()) : replacement has 0 rows, data has 204. Please help fix this.
### I used information from the output of this prompt to fix Variable 5 as well. 

# Variable 5: In_House_CDC_ELISA_S_T_Values_Log10
covid_synthetic$In_House_CDC_ELISA_S_T_Values_Log10 <- unlist(sapply(covid_synthetic$cdc, function(reactivity) {
  if (reactivity == "Reactive") {
    rnorm(1, mean = 0.4762291, sd = 0.1244573) # Mean and sd for Reactive (calculated above)
  } else {
    rnorm(1, mean = -0.7327059, sd = 0.3102606) # Mean and sd for Non-reactive (calculated above)
  }
}))
```

## Assessing Correlations from the Synthetic Dataset

I will create a correlation matrix to visualize the relationships between the variables in the original dataset. 
```{r}
# Using numerical values for the categorical variables
covid_numeric_synthetic <- covid_synthetic %>%
  mutate(
    pcr = ifelse(`pcr` == "Pos", 1, 0),
    abbott = ifelse(`abbott` == "Reactive", 1, 0),
    cdc = ifelse(`cdc` == "Reactive", 1, 0)
  ) %>%
  select(pcr, abbott, cdc)

# Compute correlation matrix
cor_matrix_synthetic <- cor(covid_numeric, method = "pearson")  
print(cor_matrix_synthetic)

# Visualize correlation matrix
library(corrplot)
corrplot(cor_matrix_synthetic, method = "color", col = colorRampPalette(c("red", "white", "blue"))(200),
         addCoef.col = "black", tl.col = "black", tl.cex = 1, number.cex = 1)
``` 
The correlation matrix above shows that the correlation coefficients dervied from the synthetic dataset variables appear to be identical to those of the original dataset. This suggests that the synthetic dataset is a good representation of the original dataset.

## Exploratory tables - categorical variables

Now, I will use Rayleen's code to explore the synthetic dataset's categorical variables.  
```{r}
# Frequency of PCR status
proc_freq(data = covid_synthetic, tables = `pcr`)

# Frequency of Abbott Reactivity
proc_freq(data = covid_synthetic, tables = `abbott`)

# Frequency of In-House CDC ELISA Reactivity
proc_freq(data = covid_synthetic, tables = `cdc`)

#Cross tab of each of the categorical variables
proc_freq(data = renames, tables = v(pcr * cdc, pcr * abbott, cdc * abbott))

```

## Exploratory anaylses - continuous variables

Now, I will use Rayleen's code to explore the synthetic dataset's continuous variables.  
```{r}
#Producing a table showing the sample size, mean, standard deviation, min, and max for each continuous variable
#Also creating a figure of each variable

#Overall stats - Abbott
proc_means(data = covid_synthetic, `Abbott_S_C_Values_Log10`)

#Stats by seroreactivity status - Abbott
proc_means(data = covid_synthetic, `Abbott_S_C_Values_Log10`, by = `abbott`)

ggplot(covid_synthetic, aes(x = `Abbott_S_C_Values_Log10`, fill = `abbott`, color = `abbott`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Non-reactive" = "maroon", "Reactive" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")


#Overall stats - CDC
proc_means(data = covid_synthetic, `In_House_CDC_ELISA_S_T_Values_Log10`)

#Stats by seroreactivity status - CDC
proc_means(data = covid_synthetic, `In_House_CDC_ELISA_S_T_Values_Log10`, by = `cdc`)

ggplot(covid_synthetic, aes(x = `In_House_CDC_ELISA_S_T_Values_Log10`, fill = `cdc`, color = `cdc`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Non-reactive" = "maroon", "Reactive" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")

#Distribution of log titers by PCR status
#Abbott
#Stats by seroreactivity status - Abbott
proc_means(data = covid_synthetic, `Abbott_S_C_Values_Log10`, by = `pcr`)

ggplot(covid_synthetic, aes(x = `Abbott_S_C_Values_Log10`, fill = `pcr`, color = `pcr`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Neg" = "maroon", "Pos" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")
#CDC
proc_means(data = covid_synthetic, `In_House_CDC_ELISA_S_T_Values_Log10`, by = `pcr`)

ggplot(covid_synthetic, aes(x = `In_House_CDC_ELISA_S_T_Values_Log10`, fill = `pcr`, color = `pcr`)) +
  geom_histogram(aes(y = ..density..), bins = 50) +
  geom_density(alpha = 0.5) +
  scale_color_manual(values = c("Neg" = "maroon", "Pos" = "black")) +
  labs(title = "Histogram with Kernel Density Overlay",
       x = "Value",
       y = "Density")
          
```

## Comparison of the Original and Synthetic datasets

I will now compare the original and synthetic datasets. I ran the same exploratory analyses that Rayleen did for optimal comparison of the two datasets. The exploratory tables for the categorical variables are similar between the datasets. The counts and PCTs between each table from the original and synthetic datasets are very similar. The exploratory analyses for the continuous variables are also similar. The first two graphs created from the synthetic dataset are similar to those from the original dataset. Their distribution is similar in that the reactive group has a higher mean value and is much more steep than the non-reactive group. However, for the last two graphs, the distribution is different. The synthetic dataset graphs are bimodal and the reactive and non-reactive groups seem to overlap. There is minimal overlap present in the original graphs. The correlation matrices for the variables appear to be identical between the original and the synthetic datasets. 
